Bayesian Data Analysis, Rough Draft

### 2023-12-01 Lahman Analysis/Formal Analysis/Draft1.qmd
Rough Draft (mostly) done
Did as much as could with the rough draft before 12/1 meeting

---
title: "Bayesian Data Analysis - Rough Draft"
author: "Sam Turner"
date: today
format: 
  html:
    toc: true
    toc-location: left
    toc-expand: true
    embed-resources: true

@@

library(tidyverse)
library(rstan)
library(Lahman)
library(bayesplot)
library(tidybayes)
library(corrplot)
```

@@

# Introduction and Motivations

@@

## Baseball Background

@@

## Literature Review

@@

# Model

@@

In this model, we say that the $HR$'s of player $n$ are binomialy distributed with number of trials for player $n$ are $AB$. The probability of success (hitting a HR) denoted as $\pi$. Furthermore, we measure $\pi$ on the logistic scale. It is obvious from the model that $\pi$ changes due to a variety of factors. The first parameter which affects $\pi$ is $\alpha_n$.

First we must noticed in the data that most hitters hit a HR in about 3 out of every 100 AB's, which can be seen in both @fig-HR.By.Age.20to40, @fig-pDist, and @fig-pDistSmall. Thus, we assume that the underlying probability of any player in the MLB for hitting a HR is about -3.5 on the logit scale, or 0.03 on the standard probability scale.

We recognize that this depends on the hitter which is where the use of a multilevel model becomes useful. Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we are able to allow our model choose an intercept which agrees with the data but can adapt depending on the player. Thus, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed).

$\beta_h$ the effect that a player being age $h$ has on HR probability. Similarly $\eta_h$ is the effect that a player's age squared has on the probability of a HR in an AB. This squared term captures the non-linearity of HR as player's age.

$\delta_p$ is the effect that a player's home ballpark $p$ has on his underlying probability of hitting a home run.

$\xi_i$ represents the change in the underlying probability of player hitting a HR in the year $i$

@@

# Results

@@

```{r}
plot(fit.poly, pars="xi")
```

# Figures

@@

>>>>

---
title: "Bayesian Data Analysis, Rough Draft"
author: "Sam Turner"
date: today
format: 
  html:
    toc: true
    toc-location: left
    toc-expand: true
    embed-resources: true

@@

library(tidyverse)
library(rstan)
library(Lahman)
library(bayesplot)
library(tidybayes)
library(corrplot)
library(boot)
```

@@

# Introduction and Motivations {#sec-intro}

@@

## Baseball Background {#sec-BaseBack}

@@

## Literature Review {#sec-LitReview}

@@

# Model {#sec-Model}

@@

In this model, we say that the $HR$'s of player $n$ follow a binomial distribution with number of trials for player $n$ in season $i$ are $AB_{ni}$. The probability of success (hitting a HR) denoted as $\pi_{ni}$ for player $n$ in season $i$. Furthermore, we measure $\pi$ on the logistic scale.

As discussed in @sec-BaseBack, there are many factors that influence player $n$'s underlying probability of hitting a $HR$ in a given $AB$ during season $i$. The first parameter which affects $\pi$ is $\alpha_n$. In our model $\alpha_n$ can be generally thought of as the term which is used to account for most player's innate ability to hit HR's. This comes from our exploratory observations of the data. In @fig-HR.By.Age.20to40, @fig-pDist, and @fit-pDistSmall, we noticed that the average (median or mean) proportion of AB's which resulted in HR's to be about 0.03. This means that about 3 in every 100 AB's result in HR's. On the logit scale, which is what our model relies on, this is about -3.5. However, we recognize this depends on the hitter which is where multilevel modeling becomes useful. Instead of simply using the average probability as our intercept, which is what a dummy variable would do in a GLM, we are able to allow our model choose an intercept which agrees with the data but can adapt depending on the player. Thus, we let $\alpha$ be normally distributed with mean $\mu_0$ (where $\mu_0$ is normally distributed with mean -3.5 and standard deviation 0.1) and standard deviation $\sigma_0$ (where the parameter $\sigma_0$ itself is exponentially distributed). By allowing $\mu_0$ and $\sigma_0$ to have their own distributions, we are able to have the distribution of $\alpha_n$ adapt to the player's specific underlying probability of hitting HR's. That is, we are able to consider what can be most closely considered their innate ability because this estimate for $\alpha_n$ does not account for other considered factors, such as age, park factors, or year played.

The next parameter we considered which affects $\pi_{ni}$ was $\beta_h$ which represents the effect age $h$, on the centered scale, has on a player's ability to hit HR's. This topic was briefly touched upon in @sec-BaseBack, where we discussed how player age can impact their their strategy and physical prowess. These ideas are well demonstrated in @fig-HR.By.Age.20to40. In @fig-HR.By.Age.20to40, one can easily see that, in general, players ability to hit HR's improves over the course of their career. As a note, we centered the the age of the all players at 30 because not only was that roughly the mean and the median, but also because it makes interpreting the $\beta_h$ easier. The value of $\beta_h$ represents how changes away from the centered age affect player $n$'s probability of hitting a HR, $\pi_{ni}$. In normal GLM's, $\beta_h$ would be general to age, meaning that there would be one intercept for each age that fit all players in the data the best. However, as with $\alpha_n$, we have employed a multilevel model to allow for each $\beta_h$ to be specific to each player. It becomes more difficult to interpret the values of $\beta_h$ because they correspond to the multiplicative effect of the centered age index. That is, in calculating the value of $\beta_h$, our model considers the priors we have set with $\mu_1$ and $\sigma_1$ to create a distribution for $\beta_h$ and then multiplies this $\beta$ with our centered age value $h$. In essence, this means that our $\beta_h$ describes the multiplicative effect of being a certain distance from the centered age, and how that effects player's $\pi$. Another important note, is that although this effect of $\beta$ is multiplicative with respect to the player's index, $h$, it is additive with respect to $\alpha$ and therefore $\pi$. This means that the whatever the model predicts for $\beta_h$ and its weight, it must be added to $\alpha_n$ to generate the correct $\pi$.

Similarly $\eta_h$ is the effect that a player's age squared has on the probability of a HR in an AB. The purpose of this squared term is to capture the non-linearity of HR as player's age. It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a non-linear effect of age on a player's ability to hit HR's. This ability appears to peak about when the player is 30-years-old. We include $\eta_h$ in the model to capture this downward trend that we see after a player reaches 30-years-old. It is nearly impossible to interpret what $\eta_h$ is doing because it is the multiplicative effect of the player's centered age squared. This begs the question, what even is a squared age? How does one interpret it? There is not a clear answer on this question, so it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with what the actual estimates are. However, it is easy to justify our choice for the priors on $\eta_h$, $\mu_2$ and $\sigma_2$. We chose $\mu_2\sim Normal(0,0.01)$ because we assumed that there would be a negative effect so small that it would be very close to zero. Further, any distinction to the slightly negative of zero would be hard to justify and, in essence, arbitrary. Therefore, we went with the mean of 0 and the standard deviation of 0.01 to account for any players who have a more or less extreme non-linearity to their trajectories. That is, higher or lower peaks and troughs throughout their career for their probability $\pi$. We chose $\sigma_2\sim Exponential(100)$ because standard deviations cannot be negative, which the exponential distribution lends itself nicely to. Furthermore, the exponential distribution with a parameter of 100 is very concentrated. Since we know that the coefficient of $\eta_h$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100). Thus, we can be confident that our standard deviation will end up being small for our estimate.

Another parameter to consider the the effect that playing in certain ballparks has on HR probability $\pi$. This was discussed in @sec-BaseBack, where we highlighted that players in certain ballparks are more likely to hit HR's than in others because of the different dimensions of the outfield fences. Furthermore, we know that these parks will effect players on different teams differently because of how schedules are created. Players will play half of their games at their home-stadium and the other half at various other ballparks known as "on the road." Thus, the home stadium that player plays at is an important factor for their HR probability $\pi$. In our model, we have estimated the effect that each ballpark has on player's playing in that ballpark. Stadiums like Great American Ballpark (index=33) tend to have a better HR effect than stadiums such as Cleveland Stadium (index=19). We account for this difference in our model with the parameter $\delta_p$ where $\delta$ is the effect that park $p$ has on a player's HR probability $\pi$. Interpreting $\delta_p$ is much easier than the previous two parameters. It simply states, on the logit scale, how the park effects the underlying probability of hitting a HR, $\pi$. Similar to $\alpha_n$ this effect is additive and can act almost like an intercept in the model. It also takes advantage of multilevel modeling since we have set priors which can adapt to the data on it, $\mu_5$ and $\sigma_5$. We have purposely set the parameter for $\mu_5\sim Normal(0,0.01)$ with mean 0 because we assume that parks in the aggregate will not have an effect on a player's ability to hit HR's. That is, for every Cleveland Stadium there exists a Great American Ballpark to counteract this. There is no formal justification for this, it is just an assumption that we make based on our preliminary exploration and our understanding that ownership (who controls the dimensions of the stadium) are not particularly biased to any ballpark size which means that we should see good variation between parks. Further, we assume that a fair standard deviation for $\mu_5$ is 0.01 because on the logistic scale, this means that we could see about a two percent change in the probabilities between any given stadiums. At most we would see a 4% difference. We chose a standard deviation for $\delta_p$ of $\sigma_5\sim Exponential(10)$ for similar reasons to all of the other $\sigma$'s. The exponential distribution cannot be negative, which we need for standard deviations. Further, we know that the effect of each park will probably not be huge because, although there is deviation, there is not extreme deviation between parks that makes some "unfair." Thus, with our parameter of 10 for the exponential distribution we are being fairly confident with our model about our estimates of the priors.

Our final parameter, $\xi_i$ represents the change in the underlying probability of player hitting a HR in the year $i$. This stems mostly from the observation in @fig-HRByYear which shows a steep increase in the amount of HR's hit per year across any subset of years considered. This means that there is an underlying change in the $\pi$ depending on the year considered. The reason for this change is irrelevant. It could be because batters are becoming better or pitchers are getting worse, or some combination of the two. Regardless, we can see that there is an increase and we need to account for this in our estimates for $\pi$ of player $n$. Thus, we include $\xi_i$ to represent the effect that playing in a certain year will have on the player's ability to hit a HR. We set the mean of $\xi_i$ as $\mu_6\sim Normal(0, 0.25)$ because we can see that the HR proportion change year over year in @fig-pByYear by only fractions of a percent. Thus, when we set our prior with mean 0 and standard deviation of 0.25, we are saying that we expect most year of year variation to be from \[-4, -3\] on the logit scale. On the normal this is \[0.018, 0.047\] which means that we would expect at most a change of 3% year over year. This is much larger than we would actually anticipate in a change but it allows the model to learn from the data to learn from the data more since we are unsure of exactly what the change beyond knowing that there is a change year over year.

@@

# Results {#sec-Results}

@@

```{r}
#| eval: false
plot(fit.poly, pars="delta")
```

```{r}
#| eval: false
print(fit.poly, pars="delta")
```

# Figures {#sec-Figures}

@@

```{r}
HomeRuns.By.Year <- Lahman::Teams %>% 
  group_by(yearID) %>% 
  summarize(HR = sum(HR),
            AB = sum(AB)) %>% 
  mutate(p = HR / AB)
```

```{r}
#| label: fig-HRByYear
#| fig-cap: "Cumulative Home Runs in MLB by Year"
ggplot(data = HomeRuns.By.Year)+
  aes(x=yearID, y=HR)+
  geom_line()+
  scale_x_continuous(n.breaks = 20)+
  labs(title = "Cumulative Home Runs in MLB by Year",
       x = "Year",
       y = "Home Runs - League Cumulative")
```

```{r}
#| label: fig-pByYear
#| fig-cap: "Cumulative Home Run Probability in MLB by Year"
ggplot(data = HomeRuns.By.Year)+
  aes(x=yearID, y=p)+
  geom_line()+
  scale_x_continuous(n.breaks = 20)+
  labs(title = "Cumulative Home Run Probability in MLB by Year",
       x = "Year",
       y = "HR Proportion (HR/AB) - League Cumulative")
```
